---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The "activity.csv" file is located in the same folder as this Rmd file, so no path is specified beyond the file name. 


```{r,echo = TRUE, results='hide'}
unzippeddatafile <- unzip("activity.zip")
data <- read.csv(file = unzippeddatafile)
file.remove(unzippeddatafile)
```


## What is mean total number of steps taken per day?

```{r}
dailysums <- tapply(data$steps, data$date, sum, na.rm = TRUE)
hist(dailysums, breaks = 50)
print(paste("The mean total number of steps taken per day is", round(mean(dailysums), 1)))
print(paste("The median total number of steps taken per day is", median(dailysums)))
```
Alternatively, the better looking way to do this (but not echo code) is:

The mean total number of steps taken per day is `r round(mean(dailysums), 1)`  
The median total number of steps taken per day is `r median(dailysums)`


## What is the average daily activity pattern?

```{r}

```


## Imputing missing values

```{r}

```


## Are there differences in activity patterns between weekdays and weekends?

```{r}

```
